Efficient ancestry and mutation simulation with msprime 1.0

### 2021-11-27 paper.tex
Update paper.tex
line 1986:  'in which no more than one group of a random number of ... '

Multiple merger coalescents, in which a random number of ancestral
lineages may merge into a common ancestor at a given time,
are referred to as $\Lambda$-coalescents.
The rate at which  a given group of $k$ out of a total of  $b$ lineages merges  is
\begin{equation}\label{lambdabk}
\lambda_{b, k} =  \int_0^1  x^{k-2}(1-x)^{b-k}\Lambda(dx) + a\one{k=2}, \quad 2 \le k \le b,
\end{equation}
where $\one{A} := 1$ if $A$ holds, and zero otherwise,
$a \geq 0$ is a constant, and $\Lambda$ is a finite measure on the
unit interval without an atom at zero \citep{DK99,P99,S99}.
There is also a larger class of simultaneous multiple merger coalescents involving
simultaneous mergers of distinct groups of lineages into several common ancestors \citep{S00}.
These are commonly referred to as $\Xi$-coalescents, and
often arise from population models incorporating diploidy or more general polyploidy
\citep{BBE13,blath2016site}.
To describe a general $\Xi$-coalescent, let $\Delta$ denote the
infinite simplex
\begin{equation*}
 \Delta := \Bigg\{ (x_1,x_2,  \ldots ): x_1 \ge x_2 \ge \cdots \ge 0, \sum_{j = 1}^{ \infty} x_j \le 1\Bigg\}.
 \end{equation*}
The rate of mergers is determined by
$\Xi = \Xi_0 + a\delta_0$, where $a \geq 0$ is a constant, $\delta_0$ is the Dirac delta
measure, and $\Xi_0$ is a finite measure on $\Delta$ with no atom at (0, 0, \ldots).
For an initial number of blocks $b \geq 2$ and
$r \in \{1,2, \ldots, b - 1 \}$, let $k_1 \geq 2, \ldots, k_r \ge 2$ be the sizes of $r$ merger events
and $s = b - k_1 - \cdots - k_r$ be the number of blocks not participating in any merger.
The rate of each possible set of mergers with sizes $(k_1, \ldots, k_r)$ is
\begin{align*}
  \lambda_{n; k_1, \ldots, k_r; s}  = {}& \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
  & +  a\one{r=1, k_1 = 2},
\end{align*}
and the number of such $(k_1, \ldots, k_r)$ mergers is
\begin{equation*}
    \mathcal{N}(b; k_1, \ldots, k_r ) = \binom{b}{k_1 \ldots k_r\, s} \frac{1}{ \prod_{j=2}^b\ell_j!  },
\end{equation*}
where $\ell_j := \#\{ i \in \{ 1, \ldots, r \} : k_i = j \}$ is the number of mergers of size $j \geq 2$ \citep{S00}.

>>>>

Multiple merger coalescents, in which no more than one group of a random number of ancestral
lineages may merge into a common ancestor at a given time,
are referred to as $\Lambda$-coalescents.
The rate at which  a given group of $k$ out of a total of  $b$ lineages merges  is
\begin{equation}\label{lambdabk}
\lambda_{b, k} =  \int_0^1  x^{k-2}(1-x)^{b-k}\Lambda(dx) + a\one{k=2}, \quad 2 \le k \le b,
\end{equation}
where $\one{A} := 1$ if $A$ holds, and zero otherwise,
$a \geq 0$ is a constant, and $\Lambda$ is a finite measure on the
unit interval without an atom at zero \citep{DK99,P99,S99}.
There is also a larger class of simultaneous multiple merger coalescents involving
simultaneous mergers of distinct groups of lineages into several common ancestors \citep{S00}.
These are commonly referred to as $\Xi$-coalescents, and
often arise from population models incorporating diploidy or more general polyploidy
\citep{BBE13,blath2016site}.
To describe a general $\Xi$-coalescent, let $\Delta$ denote the
infinite simplex
\begin{equation*}
 \Delta := \Bigg\{ (x_1,x_2,  \ldots ): x_1 \ge x_2 \ge \cdots \ge 0, \sum_{j = 1}^{ \infty} x_j \le 1\Bigg\}.
 \end{equation*}
The rate of mergers is determined by
$\Xi = \Xi_0 + a\delta_0$, where $a \geq 0$ is a constant, $\delta_0$ is the Dirac delta
measure, and $\Xi_0$ is a finite measure on $\Delta$ with no atom at (0, 0, \ldots).
For an initial number of blocks $b \geq 2$ and
$r \in \{1,2, \ldots, b - 1 \}$, let $k_1 \geq 2, \ldots, k_r \ge 2$ be the sizes of $r$ merger events
and $s = b - k_1 - \cdots - k_r$ be the number of blocks not participating in any merger.
The rate of each possible set of mergers with sizes $(k_1, \ldots, k_r)$ is
\begin{align*}
  \lambda_{n; k_1, \ldots, k_r; s}  = {}& \int_\Delta  \sum_{\ell = 0}^s \sum_{\substack {i_1, \ldots, i_{r+\ell} = 1\\ \text{all distinct}} }^\infty  \binom{s}{\ell} x_{i_1}^{k_1} \cdots  x
_{i_{r}}^{k_r} x_{i_{r+1}} \cdots x_{i_{r+\ell}}\left(1 - \sum_{j=1}^\infty x_j \right)^{s-\ell} \frac{1}{ \sum_{j=1}^\infty x_j^2 } \Xi_0(dx)   \\
  & +  a\one{r=1, k_1 = 2},
\end{align*}
and the number of such $(k_1, \ldots, k_r)$ mergers is
\begin{equation*}
    \mathcal{N}(b; k_1, \ldots, k_r ) = \binom{b}{k_1 \ldots k_r\, s} \frac{1}{ \prod_{j=2}^b\ell_j!  },
\end{equation*}
where $\ell_j := \#\{ i \in \{ 1, \ldots, r \} : k_i = j \}$ is the number of mergers of size $j \geq 2$ \citep{S00}.
